---
title: "Stt864 Lab4"
author: "Nan Cao"
date: "April 27, 2016"
output: pdf_document
---

Data preparing
```{r}
setwd("C://Users//nan66//Google Drive//stt864//LAB4")
load(file="Alldata.Rdata")
set.seed(52871775)
library(nlme)
library(lme4)
library(MASS)
library(glmmML)
```
```{r,echo=FALSE}
load(file="GOTermlist.RData")
load(file="pvalset.Rdata")
load(file="pvalset2.Rdata")
```
  
Q1
```{r}
Goterms<-Alldata$originaldata[,128:130]
mc<-dim(Goterms)[1]
```
```{r,eval = FALSE}
GOTermlist<-NULL
for (j in 1:mc)
{
 	list<-NULL
 	for (k in 1:3)
	     {
		GOTerm22283<-as.character(Goterms[j,k])
		getGoTerms<-unlist(strsplit(GOTerm22283,"///"))
		repj<-rep(j,length(getGoTerms))	
		newlist<-cbind(repj,getGoTerms)
		list<-rbind(list,newlist)	
	     }
 	GOTermlist<-rbind(GOTermlist,list)	
}	
```

```{r}
GO0006955<-which(GOTermlist[,2]=="GO:0006955")
rownums<-as.numeric(GOTermlist[GO0006955,1])
subsetGenes<-Alldata$originaldata[rownums,c(3:110)]
write.table(subsetGenes,file="subsetGenes.txt")
```
  
Q2
```{r}
samIDs<-names(subsetGenes)
Beverages<-((samIDs%in%Alldata$trt1)*1
           +(samIDs%in%Alldata$trt2)*2
           +(samIDs%in%Alldata$trt3)*3
           +(samIDs%in%Alldata$trt4)*4)
Subject<-((samIDs%in%Alldata$ind1)*1
         +(samIDs%in%Alldata$ind2)*2
         +(samIDs%in%Alldata$ind3)*3
         +(samIDs%in%Alldata$ind4)*4
         +(samIDs%in%Alldata$ind5)*5
         +(samIDs%in%Alldata$ind6)*6)
hours<-((samIDs%in%Alldata$time_h0)*0
       +(samIDs%in%Alldata$time_h1)*1
       +(samIDs%in%Alldata$time_h2)*2
       +(samIDs%in%Alldata$time_h4)*4
       +(samIDs%in%Alldata$time_h12)*12)
```
  
Q3
```{r}
BeverFac<-as.factor(Beverages)
hourFac<-as.factor(hours)
resp<-as.numeric(subsetGenes[1,])
lmmd2<-lmer(resp~BeverFac+(1|Subject)+(1|hours))
summary(lmmd2)
```

REML estimates of variances:
```{r}
lmreg<-lm(resp~BeverFac+Subject+hours)
anova(lmreg)
summary(lmreg)
sigmahat<-sqrt(0.07058)
sigmahat
sigmahat_Subject<-(0.15870-0.07058)/1
sigmahat_Subject
sigmahat_hours<-(0.34036-0.07058)/1
sigmahat_hours
```
$H_0$ : the means are the same; $H_1$ : the means are different.
Statistical model:
\begin{equation*}
Z=\frac{ \hat{\beta_1} - \hat{\beta_2} -1}
{\sqrt{Var(\hat{\beta_1})+Var(\hat{\beta_2})-2Cov(\hat{\beta_1},\hat{\beta_2})}}\sim Z
\end{equation*}
For test the significance of difference between alcohol and water group:
```{r}
summary(lmmd2)$coefficients[4,3]
2*(1-pnorm(abs(summary(lmmd2)$coefficients[4,3])))
```
Reject the null hypothesis, the means are significantly different.  
  
For test the significance of difference between alcohol and water group:
```{r}
Zstat2<-(summary(lmmd2)$coefficients[3,1]
         -summary(lmmd2)$coefficients[4,1])/sqrt(vcov(lmmd2)[3,3]
         +vcov(lmmd2)[4,4]-2*vcov(lmmd2)[4,3])
Zstat2
pnorm(abs(Zstat2))
```
We can't reject the null hypothesis.  
  
Q4  
```{r}
numgenes<-dim(subsetGenes)[1]
```
```{r,eval=FALSE}
pvalset<-rep(0,numgenes)
pvalset2<-rep(0,numgenes)
for (i in 1:numgenes)
{
 resp<-as.numeric(subsetGenes[i,])
 lmmd2<-lmer(resp~BeverFac+(1|Subject)+(1|hours))
 Zstat<-summary(lmmd2)$coefficients[4,3]
 Zstat2<-(summary(lmmd2)$coefficients[3,1]
          -summary(lmmd2)$coefficients[4,1])/sqrt(vcov(lmmd2)[3,3]
          +vcov(lmmd2)[4,4]-2*vcov(lmmd2)[4,3])
 pvalset[i]<-2*(1-pnorm(abs(Zstat)))
 pvalset2[i]<-2*(1-pnorm(abs(Zstat)))
}
```

```{r,fig.width = 8, fig.height = 4}
smallest<-which.min(pvalset)
smallest
smallest2<-which.min(pvalset2)
smallest2
oldpar <- par(mfrow = c(1, 2))
hist(pvalset,main="p-value of hypothesis 1")
hist(pvalset2,main="p-value of hypothesis 2")
par(oldpar)
0.05/numgenes
min(pvalset)
min(pvalset2)
```
Since there's no gene significant for the second hypothesis, there's no gene significant for both hypothesises.
```{r}
newresp<-as.numeric(subsetGenes[smallest,])
lmmd3<-lmer(newresp~BeverFac+(1|Subject)+(1|hours))
summary(lmmd3)
newresp2<-as.numeric(subsetGenes[smallest2,])
lmmd32<-lmer(newresp2~BeverFac+(1|Subject)+(1|hours))
summary(lmmd32)

```
Q5
```{r}
plot(hours,newresp,type="n",xlab="Hours", ylab="gene expression")
for (sub in 1:6)
 for (bever in 1:4)
{
points(
      hours[(Subject==sub)&(Beverages==bever)],
      newresp[(Subject==sub)&(Beverages==bever)],
      col=bever,pch=sub
      )
lines(
      hours[(Subject==sub)&(Beverages==bever)],
      newresp[(Subject==sub)&(Beverages==bever)],
      col=bever
      )
}
```
There're obvious changes of gene expression in the first 2 hours, but the change from the 2nd hour to the 12th hour.  

Q6
```{r}
resp<-newresp
hours2<-(hours^2)
hours3<-(hours^3)
combfacs<-Subject*10+Beverages
glscsh<-gls(
	resp~BeverFac+hours+hours2+hours3,
	correlation=corCompSymm(form=~1|combfacs),
	weights=varIdent(form=~1|combfacs),
	method="REML"
	)
summary(glscsh)
glsarh1<-gls(
	resp~BeverFac+hours+hours2+hours3,
	correlation=corAR1(form=~1|combfacs),
	weights=varIdent(form=~1|combfacs),
	method="REML"
	)
summary(glsarh1)
glscsM2<-gls(
	resp~BeverFac+BeverFac*hours+BeverFac*hours2+BeverFac*hours3,
	correlation=corCompSymm(form=~1|combfacs),
	method="REML"
	)
summary(glscsM2)
glscshM2<-gls(
	resp~BeverFac+BeverFac*hours+BeverFac*hours2+BeverFac*hours3,
	correlation=corCompSymm(form=~1|combfacs),
	weights=varIdent(form=~1|combfacs),
	method="REML"
	)
summary(glscshM2)
```
According to the summary of the fitted model ,the last two model seem better.
The time is significant only in the last models.  

Q7  
Define new data set
```{r}
Combdata<-cbind(newresp,hours,Subject,Beverages,combfacs)
unicombfacs<-Combdata[Combdata[,2]==0,5]
newcombdata<-NULL
for (i in 1:length(unicombfacs))
{
 subcomb<-Combdata[Combdata[,5]==unicombfacs[i],]
 Yvec<-subcomb[,1]
 Hvec<-subcomb[,2]
 Zvec<-((Yvec[Hvec!=0]-Yvec[Hvec==0])>0)+0
 newcombdata<-rbind(newcombdata,cbind(Zvec,subcomb[Hvec!=0,c(2:5)]))
}

newcombdata1<-list(
                   Zvec=newcombdata[,1],
                   hours=newcombdata[,2],
                   Subject=newcombdata[,3],
                   Beverages=newcombdata[,4],
                   combfacs=newcombdata[,5]
                   )
newcombdata1$Beverages<-as.factor(newcombdata1$Beverages)
newcombdata1$hours<-as.factor(newcombdata1$hours)
```
  
GLMM with the default Laplace approximation
```{r}
glmRandomIntcept1<-glmer(
   newcombdata1$Zvec~newcombdata1$Beverages+newcombdata1$hours
  +(1|newcombdata1$combfacs),family=binomial
  )
summary(glmRandomIntcept1)
```
  
Another way to fit using Laplace approximation  
```{r}
glmRandomIntcept2<-glmmML(
  newcombdata1$Zvec~newcombdata1$Beverages+newcombdata1$hours,
  family=binomial,cluster=newcombdata1$combfacs)
summary(glmRandomIntcept2)
```
  
Fit the model using the penalized quasi-likelihood
```{r}
newcombdata2<-data.frame(newcombdata1)
glmRandomIntcept3<-glmmPQL(
                           Zvec~Beverages+hours,
                           random=~1|combfacs,family=binomial,
                           data=newcombdata2
                           )
summary(glmRandomIntcept3)
```
  
Fit the model using adaptive Gauss-Hermite qudrature(AGQ)  
nAGQ: the number of points per axis for evaluating the adaptive Gauss-Hermite approximation to the log-likelihood.  
```{r,fig.width = 8, fig.height = 6}
glmRandomIntcept4<- glmer(
  newcombdata1$Zvec~newcombdata1$Beverages+newcombdata1$hours+(1|newcombdata1$combfacs),
  nAGQ=8,family=binomial)
summary(glmRandomIntcept4)
oldpar <- par(mfrow = c(2, 3))
plot(residuals(glmRandomIntcept1))
plot(residuals(glmRandomIntcept3))
plot(residuals(glmRandomIntcept4))
qqnorm(residuals(glmRandomIntcept1))
abline(0,1)
qqnorm(residuals(glmRandomIntcept3))
abline(0,1)
qqnorm(residuals(glmRandomIntcept4))
abline(0,1)
par(oldpar)
```
Compared among the 3 models, only time effects are signficant in all the models.
The  penalized quasi-likelihood model abd  model using adaptive Gauss-Hermite qudrature have lower correlation of fixed effects.  
What's more, we failed to get the residuals of the second model because the infinite value.   
The first model and the forth model has the same residual plot and qqnorm-plot, they both shows higher normaility compared with the third model.
